---
title: "__Kane County ADA Transition Plan__"
author: |
  | __Credits__:
  | Troy Simpson - Kane County Division of Transportation
date: "__Updated:__ `r format(Sys.Date(), '%B %d, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r libraries, include = FALSE}
library(ymlthis)
library(knitr)
library(kableExtra)
library(ggmap)
library(shiny)
library(tableHTML)
library(prettydoc)
library(tidyr)
library(dplyr)
library(formattable)
```

# Introduction
***
## Background

### The Americans with Disabilities Act

The Americans with Disabilities Act (ADA) was signed into law on July 26, 1990, by President George H.W. Bush. The Act is a civil rights statute that prohibits discrimination against people who have disabilities. There are five separate Titles of the Act that relate to potential discrimination. Title II of the Act specifically addresses the topic of making sure public services and public transportation are accessible to those with disabilities. With the arrival of the Act, designing and constructing public facilities that are not accessible by people with disabilities constitutes discrimination.

### ADA Accessibility Guidelines: Limitations for Transportation

However, ADA standards for new construction and alterations were primarily developed for buildings and site work and were not easily applicable to sidewalks, street crossings, and related pedestrian facilities in the public right-of-way. Similarly, Section 504 standards did not offer guidance appropriate for right-of-way construction. The need to address right-of-way accessibility in a more specific way became apparent from the difficulties agencies had in applying ADA Accessibility Guidelines (ADAAG). In 1999, the United States Access Board, the Federal Government’s specialist in accessible design, stated the rule making process for accessible pedestrian facilities in public right-of-way by convening a Federal advisory committee of key stakeholders to develop recommendations that could supplement or replace the current standard. The Public Rights-Of-Way Access Advisory Committee (PROWAAC) completed its initial work in 2000 and published its recommendations for new guidelines in a report to the Transportation Research Board in 2001.

### Public Rights-of-Way Access Guidelines (PROWAG)

In 2002, the Access Board issued a Notice of Availability of Draft Public Rights-of-Way Accessibility Guidelines (PROWAG) based on the PROWAAC report. Comments from consumers and design professionals led to the issuance of a second draft in 2005. A Notice of Proposed Rule Making (NPRM) seeking public comment prior to publication of a final rule was sent out in 2013. When standards consistent with the final PROWAG guidelines are adopted by the Department of Justice they will become the new minimum design standards under the ADA for both new construction and alterations. \textcolor{red}{Insert a section about Illinois Accessibility Code/IDOT Requirements for Public Agencies.}

### Public Agency ADA Compliance: Maintenance vs. Alteration

The Act applies to all public facilities built before and after the arrival of the Act. The ADA guidelines apply to newly constructed facilities and altered portions of existing facilities. However, the ADA guidelines only include existing facilities if the scope of the improvement includes alteration. *__Alteration__* includes reconstruction, rehabilitation, resurfacing, widening, and projects of similar scale and effect. *__Maintenance__* activity does not constitute an alteration under the ADA guidelines. _Below is a guide for determining if an activity triggers an ADA audit of the facilities_.

```{r maintenance-alteration-table}

#Create two vectors for maintenance and alteration work types

maintenance <- c('Chip Seals', 'Fog Seals', 'Scrub Sealing', 'Crack Filling and Sealing', 'Joint Crack Seals', 'Slurry Seals', 'Diamond Grinding', 'Joint Repairs', 'Spot High-Friction Treatments', 'Dowel Bar Retrofit', 'Pavement Patching', 'Surface Sealing')
alteration <- c('Addition of New Layer of Asphalt', 'Mill & Fill/Mill and Overlay', 'Cape Seals', 'New Construction', 'Hot-in-Place Recycling','Open-Graded Surface Course', 'Micro-surfacing/Thin-Lift Overlay', 'Rehabilitation and Reconstruction')

#Create vectors for maintenance and alteration attributes, ADA audit required attributes

maintenance_attr <- rep("Maintenance", length(maintenance))
alteration_attr <- rep("Alteration", length(alteration))
audit_required_no <- rep("No", length(maintenance))
audit_required_yes <- rep("Yes", length(alteration))

#Combine vectors

work_type <- c(maintenance, alteration)
work_type_adaclass <- c(maintenance_attr, alteration_attr)
ada_audit_required <- c(audit_required_no, audit_required_yes)

#Combine into data frame and order by alphabet and ADA work type

maintenance_alteration_df <- data.frame(cbind(work_type, work_type_adaclass, ada_audit_required)) %>% arrange(work_type_adaclass, work_type)

maintenance_alteration_df %>% 
  mutate(
    ada_audit_required = 
      cell_spec(ada_audit_required,
                color = ifelse(ada_audit_required == "Yes", "#1a651a", "#e60000"))
  ) %>%
  kable(format = "pipe", col.names = c("Work Type", "ADA Classification", "ADA Facilities Audit Required?"), caption = "Maintenance vs. Alteration Activities")
```
## Applicability to the Kane County Division of Transportation

As a necessary step to a program access plan to provide accessibility under the Act, the Kane County Division of Transportation is required to perform a self-evaluation of the current facilities relative to the accessibility requirements of the Act. The Kane County DOT is then required to develop an ADA Transition Plan to address any deficiencies in public facilities. As facilities are updated to current ADA standards, the plan is required to be updated periodically until all barriers are removed. The Plan is intended to achieve the following:

1. Identify physical obstacles that limit the accessibility of facilities to individuals with disabilities
2. Describe the methods to be used to make the facilities accessible
3. Provide a schedule for making the access modifications
4. Identify the public officials responsible for implementation of the Transition Plan

## 7 Steps to ADA Compliance

- Step 1 - Designate an ADA Coordinator
- Step 2 - Provide notice to the public about ADA requirements
- Step 3 - Establish a grievance procedure
- Step 4 - Develop internal design standards, specifications, and details
- Step 5 - The ADA Transition Plan
- Step 6 - Schedule and budget for improvements
- Step 7 - Monitor the progress on the implementation of the Transition Plan

*Each step to ADA compliance is described in detail in the following section*.

# Steps to ADA Compliance

***
## Designate an ADA Coordinator

The Kane County Division of Transportation must designate at least one responsible employee to coordinate ADA compliance. The benefits of having an ADA coordinator are:

- It makes it easier for members of the public to identify someone to help them with questions and concerns about disability discrimination.
- It provides a single source of information so questions from KDOT staff and from outside KDOT can be answered quickly and consistently.
- It provides an individual who can focus on and who can be instrumental in moving compliance plans forward.

The person who is appointed to this position must be familiar with the operations of KDOT, trained in the requirements of the ADA and other laws pertaining to discrimination, and be able to deal effectively with local governments, advocacy groups, and the public. It is assumed that the coordinator is given sufficient time free of other responsibilities to carry out the functions of the position. The ADA coordinator for the Kane County Division of Transportation is currently:

\begin{center}

\textbf{Troy Simpson} \\
ADA Coordinator \\
Kane County Division of Transportation \\
41W011 Burlington Rd \\
St Charles, IL 60175 \\
(630) 444-3142 \\
\href{mailto:kdotada@co.kane.il.us}{\textcolor{blue}{kdotada@co.kane.il.us}}

\end{center}

## Provide Notice to the Public about ADA Requirements

The Kane County Division of Transportation must provide public notice about the rights of the public under the ADA and the responsibility of KDOT under the ADA. Providing notice is not a one-time requirement, but a continuing responsibility. The audience of those who may have an interest in accessibility on KDOT facilities might include a large number of individual citizens that are not easily identified. Groups that are likely to include the target audience include public transit users and advocacy groups. KDOT has the responsibility to determine the most effective way to provide notice. KDOT will provide public notice on via official website and will also reach out to impacted advocacy groups and local agencies near KDOT right-of-way. The public and selected groups for outreach will have thirty days to provide comments into the formation of this plan.

## Establish a Grievance Procedure

The Kane County Division of Transportation is required to adopt and publish procedures for resolving grievances arising under Title II of the ADA. The procedures are intended to set out a system for resolving complaints of disability discrimination in a prompt and fair manner. Complaints would typically be directed to a department's Office of Civil Rights as applicable. It is generally thought that filing a complaint with KDOT is an appropriate first step, in that it provides an opportunity to resolve a local issue at the local level; however, the exhaustion of KDOT's grievance procedure is not a prerequisite to filing a complaint with either a federal agency or court.

*A copy of the Kane County Division of Transportation's Grievance Procedure and Grievance Form can be found in Appendix F. The Grievance Procedure will also be posted on KDOT's website.*

## Develop Internal Design Standards, Specifications, and Details

The Architectural and Transportation Barrier Compliance Board (alternatively called the Access Board) has developed accessibility guidelines for pedestrian facilities in the public right-of-way. The Federal Highway Administration has recognized these as its currently recommended best practices. The Kane County Division of Transportation can adopt these accessibility guidelines into its own system of standards, specifications, and design details with modifications to meet local conditions. Development of design standards and design details within KDOT will allow for consistency in the application of ADA requirements for new facilities. The Illinois Department of Transportation has also brought the ADA standards into their own design standards. Local agencies in Illinois can use Section 41-6 of the Bureau of Local Roads and Streets Manual to meet the design standards.

At this point in time KDOT does not have any adopted design standards, but does reference the Illinois Department of Transportation's adopted standards. Should KDOT adopt design standards in the future, those standards will reference ADA design standards.

KDOT can also review and revise County ordinances and code to reference current ADA standards. For example, Kane County can save costs in the future by restricting the installation of power poles, fire hydrants, hand holes, drainage grates and other barriers in areas like to require pedestrian elements.

## The ADA Transition Plan

The ADA Transition Plan should consist of the following elements:

- A list of physical barriers in KDOT's facilities that limit accessibility of individuals with disabilities (Self-Evaluation)
- A detailed description of the methods to remove these barriers and make the facilities accessible.
- A schedule for taking the necessary steps.
- The name of the official responsible for implementation
- A schedule for providing curb ramps.
- A record of the opportunity given to the disability community and other interested parties to participate in the development of the plan.

Periodic updates to the ADA Transition Plan are required in order to ensure on-going compliance. Some of these key elements of the Transportation Plan are described in greater detail below:

### Self-Evaluation

## Schedule and Budget for Improvements
## Monitor the Progress on the Implementation of the Transition Plan
# Conclusion
***
# Appendices
***
## Appendix A - Curb Ramp Inventory
## Appendix B - Sidewalk/Trail Evaluation
## Appendix C - Bus Shelter Inventory
## Appendix D - KDOT Building ADA Assessment
## Appendix E - Public Notice and Participation
## Appendix F - Grievance Procedure



